---
title: "Exploratory Data Analysis"
format: html
editor: visual
---

# Elements of structured data

Data types:

-   Numeric: data that are expressed on a numeric scale

    -   Continuous: data that can take on any value in an interval (synonyms: interval, float, numeric)

    -   Discrete: data than can take on only integer values, such as counts (synonyms: integer, count)

-   Categorical: data that can take on only a specific set of values representing a set of possible categories (synonyms: enums, enumerated, factors, nominal)

    -   Binary: a special case of categorical data with just two categories of values, e.g. 0/1, true/false (synonyms: dichotomous, logical, indicator, boolean)

    -   Ordinal: categorical data that has an explicit ordering (synonym: ordered factor)

# Rectangular data

-   Data frame: rectangular data (like a spreadsheet) is the basic data structure for statistical and machine learning models

-   Feature: a column within a table is commonly referred to as a *feature* (synonyms: attribute, input, predictor, variable)

-   Outcome: many data science projects involve predicting an *outcome* - often a yes/no outcome. The *features* are sometimes used to predict the *outcome* in an experiment or a study (synonyms: dependent variable, response, target, output)

-   Records: a row within a table is commonly referred to as a *record* (synonyms: case, example, instance, observation, pattern, sample)

# Estimates of location

-   Mean : the sum of all values divided by the number of values (synonym: average)

    $$
    Mean=\bar{x}=\frac{\sum_{i=1}^n x_i}{n}
    $$

-   Weighted mean: the sum of all values times a weight divided by the sum of the weights (synonym: weighted average)

$$
Weighted\ mean=\bar{x}_w=\frac{\sum_{i=1}^n w_ix_i}{\sum_{i=1}^n w_i}
$$

-   Median: the value such that one-half of the data lies above and below (synonym: 50th percentile)

-   Percentile: the value such that P percent of the data lies below (synonym: quantile)

-   Weighted median: the value such that one-half of the sum of the weights lies above and below the sorted data

-   Trimmed mean: the average of all values after dropping a fixed number of extreme values (synonym: truncated mean)

$$
Trimmed\ mean=\bar{x}=\frac{\sum_{i=p+1}^{n-p}x_{(i)}}{n-2p}
$$

-   Robust: not sensitive to extreme values (synonym: resistant)

-   Outlier: a data value that is very different from most of the data (synonym: extreme value)

## Example: location estimates of population and murder rates

```{r}
state <- read.csv("data/state.csv")
head(state)
```

```{r}
mean(state[['Population']])
mean(state[['Population']], trim = 0.1)
median(state[['Population']])
```

```{r}
weighted.mean(state[['Murder.Rate']], w = state[['Population']])
```

```{r}
matrixStats::weightedMedian(state[['Murder.Rate']], w = state[['Population']])
```

# Estimates of variability

-   Deviations: the difference between the observed values and the estimate of location

    -   synonyms: errors, residuals

-   Variance: the sum of squared deviations from the mean divided by $n-1$ where $n$ is the number of data values

    -   synonym: mean-squared error

-   Standard deviation: the square root of the variance

-   Mean absolute deviation: the mean of the absolute values of the deviations from the mean

    -   synonyms: l1-norm, Manhattan norm

-   Median absolute deviation from the median: the median of the absolute values of the deviations from the median

-   Range: the difference between the largest and the smallest values in a data set

-   Order statistics: metrics based on the data values sorted from smallest to biggest

    -   synonym: ranks

-   Percentile: the value such that $P$ percent of the values take on this value or less and (100-$P$) percent take on this value or more

    -   synonym: quantile

-   Interquartile range: the difference between the 75th percentile and the 25th percentile

    -   synonym: IQR

## Standard deviation and related estimates

Mean absolute deviation :

$$
Mean\ absolute\ deviation=\frac{\sum_{i=1}^n |{x_i-\bar{x}}|}{n}
$$

Variance :

$$
Variance=s^2=\frac{\sum_{i=1}^n(x_i-\bar{x})^2}{n-1}
$$

Standard deviation :

$$
Standard deviation=s=\sqrt{variance}
$$

Neither the variance, the standard deviation, nor the mean absolute deviation are robust to outliers and extreme values.

A robust estimate of variability is the *median absolute deviation from the median* or MAD:

$$
Median\ absolute\ deviation=Median(|x_1-m|, |x_2-m|, ..., |x_N-m|)
$$

where $m$ is the median

## Estimates based on percentiles
